Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-15 write-ups/PLDI/paper.tex
Polishing implementation.

\subsection{The megakernel approach}\label{sec:megakernel}

The mock-up works as follows.  Instead of multitasking multiple
separate kernels, we merge a set of kernels into a megakernel---a
single, monolithic kernel---and encode the scheduling logic inside
this megakernel.  The megakernel is launched with as many workgroups
as can run concurrently according to the occupancy-bound.  One
workgroup takes the role of scheduler, and the remaining workgroups
act as a pool of workers.  A worker repeatedly queries the scheduler
to be assigned a task.  A task corresponds to executing a cooperative
or non-cooperative kernel.  In the case where a task corresponds to a
non-cooperative kernel, the workgroup executes the kernel function
uninterrupted and awaits further work.  In the case of a cooperative
kernel, the workgroup either starts from the kernel entry point,
corresponding to the workgroup being an \emph{initial} workgroup of
the kernel, or immediately jumps to a designated point within the
cooperative kernel, corresponding to the workgroup being forked by an
already executing workgroup.  In the latter case, the new workgroup
also receives a struct containing the values of all relevant
$\transmit$-annotated variables.

@@

For ease of implementation, our current prototype supports
multitasking a single cooperative kernel with a single non-cooperative
kernel (though the non-cooperative kernel can be invoked many times).
Rather than adding language support for $\transmit$ annotations, we
assume that all variables declared at the root lexical scope of a
cooperative kernel are $\transmit$-annotated; this allows our
implementation to piggy-back on top of Clang (see below) without
requiring front-end changes.  Finally, we require that $\offerkill$,
$\offerfork$ and $\resizingglobalbarrier$ are called from the entry
function of a cooperative kernel.  This allows us to use
$\keyword{goto}$ to direct threads joining the cooperative kernel to
the appropriate location, and $\keyword{return}$ to force threads to
leave the computation.  These restrictions still allow us to
experiment with interesting irregular algorithms (see
\mysec\ref{sec:experiments}), and could all be removed with
engineering effort, but this effort might not be well invested since a
full implementation of coopeartive kernels would not use the
megakernel approach we present here.

@@

To enable multitasking through cooperative kernels, the runtime (in
our case, the mega-kernel), must (a) track the state of all
workgroups, i.e.\ if a workgroup is waiting or computing a kernel, (b)
maintain consistent context states for each kernel, e.g.\ the number
of active workgroups computing a cooperative kernel, and (c) provide a
safe way for these states to be modified in response to $\offerfork$ and $\offerkill$. We now discuss these issues in more detail.

\paragraph{Kernel context}

When a workgroup executes a kernel (cooperative or otherwise), it must
be able to query a consistent context to obtain a workgroup id value
and the
number of workgroups executing the kernel. In single kernel execution,
the OpenCL primitives $\getnumgroups()$ and $\getgroupid()$ provide this functionality natively. For multiple
kernel execution, we emulate these primitives via a per kernel \emph{context}, containing, for a given kernel: the number of active groups $M$, the total number of requested groups $N$, and a mapping from the native workgroup id to a new workgroup
id valid for the kernel.  In the case of a non-cooperative kernel, $M$ is undefined because the notion of active workgroups is not relevant.

\paragraph{Scheduler context}

To dynamically manage workgroups executing cooperative
kernels, our framework must track the state of each workgroup and
provide a channel of communication from the scheduler to
workgroups executing $\offerfork$ and $\offerkill$. To achieve this,
we use a \emph{scheduler context} structure, mapping a primitive workgroup id the workgroup's status, which is either \emph{available}, or the id of the kernel that the workgroup is currently executing.  The scheduler can then send cooperative
kernels a \emph{resource message}, commanding workgroups executing
the kernel to exit at $\offerkill$, or 
spawn additional workgroups at $\offerfork$. Thus, the
scheduler context needs a communication channel for each cooperative
kernel. In our case, the communication channel is implemented using
global atomic variables.

\paragraph{Launching kernels}

To launch a kernel, the host sends a data packet to the GPU
consists of: a kernel to execute, the inputs to the kernel, the
number of workgroups that should execute the kernel, and a flag indicating whether the kernel
is cooperative. In our implementation, this communication channel
between host and device is achieved through fine-grained SVM atomics.

\paragraph{Dynamically managing workgroups}
On receiving a data packet from the host
describing a kernel launch $K$, the GPU scheduler must make decisions
about how to schedule $K$. Suppose $K$ requests
$N$ workgroups. The scheduler queries the scheduler context.  If
there are at least $N$ available workgroups, $K$ can be scheduled
immediately. Suppose instead that there are only $N_a < N$ available workgroups, but a cooperative kernel $K_c$ is
executing. The scheduler can use $K_c$'s channel in the scheduler context to command $K_c$ to provide $N - N_a$ workgroups via $\offerkill$.
Once $N$ workgroups are available, the scheduler initialises a
kernel context $\KC$ for $K$, setting the number of active workgroups
to $N$ and mapping available workgroups (located in the scheduler
context) to their new $\KC$ workgroup ids. The scheduler then sends $N$
workgroups from the available workgroups to execute kernel $K$ with
kernel context $\KC$.  Note that if the new kernel $K$ is itself a cooperative kernel, the scheduler could decide to provide $K$ with fewer than $N$ active workgroups initially.

>>>>

\subsection{The megakernel mock up}\label{sec:megakernel}

Instead of multitasking multiple separate kernels, we merge a set of
kernels into a megakernel---a single, monolithic kernel.  The
megakernel is launched with as many workgroups as can be occupant
concurrently.  One workgroup takes the role of scheduler, and the
scheduling logic is embedded as part of the megakernel.  The remaining
workgroups act as a pool of workers.  A worker repeatedly queries the
scheduler to be assigned a task.  A task corresponds to executing a
cooperative or non-cooperative kernel.  In the non-cooperative case,
the workgroup executes the relevant kernel function uninterrupted,
then awaits further work.  In the cooperative case, the workgroup
either starts from the kernel entry point or immediately jumps to a
designated point within the kernel, depending on whether the workgroup
is an initial workgroup of the kernel, or a forked workgroup.  In the
latter case, the new workgroup also receives a struct containing the
values of all relevant $\transmit$-annotated variables.

@@

For ease of implementation, our current prototype supports
multitasking a single cooperative kernel with a single non-cooperative
kernel (though the non-cooperative kernel can be invoked many times).
Rather than adding language support for $\transmit$ annotations, we
assume that all variables declared at the root lexical scope of a
cooperative kernel are $\transmit$-annotated; this allows our
implementation to piggy-back on top of Clang (see below) without
requiring front-end changes.  Finally, we require that $\offerkill$,
$\offerfork$ and $\resizingglobalbarrier$ are called from the entry
function of a cooperative kernel.  This allows us to use
$\keyword{goto}$ and $\keyword{return}$ to direct threads into and out
of the kernel.  With these restrictions we can experiment
with interesting irregular algorithms (see
\mysec\ref{sec:experiments}).  A non-mock implementation of
coopeartive kernels would not use the megakernel approach, so we do not deem
the engineering effort associated with lifting these restrictions to
be worthwhile.

@@

To enable multitasking through cooperative kernels, the runtime (in
our case, the mega-kernel) must track the state of all
workgroups, i.e.\ whether a workgroup is waiting or computing a kernel,
maintain consistent context states for each kernel, e.g.\ tracking the number
of active workgroups, and provide a
safe way for these states to be modified in response to $\offerfork$/$\offerkill$. We discuss these issues,
and describe the implementation of an efficient resizing barrier.

We describe how the scheduler would handle arbitrary
combinations of kernels, though as noted above our current
implementation is restricted to the case of two kernels.

\paragraph{Kernel and scheduler contexts}

When a workgroup executes a kernel (cooperative or otherwise), it must
be able to query a consistent context to obtain a workgroup id value
and the
number of workgroups executing the kernel. For multi-kernel
kernel execution, we emulate the $\getnumgroups$ and $\getgroupid$ 
primitives via a per kernel \emph{context}.  For a given kernel this records: the number of active groups $M$ (irrelevant for a non-cooperative kernel), the total number of requested groups $N$, and a mapping from the native workgroup id to a new workgroup
id valid for the kernel.

To dynamically manage workgroups executing cooperative
kernels, our framework must track the state of each workgroup and
provide a channel of communication from the scheduler workgroup to
workgroups executing $\offerfork$ and $\offerkill$. To achieve this,
we use a \emph{scheduler context} structure, mapping a primitive workgroup id the workgroup's status, which is either \emph{available} or the id of the kernel that the workgroup is currently executing.  The scheduler can then send cooperative
kernels a \emph{resource message}, commanding workgroups to exit at $\offerkill$, or 
spawn additional workgroups at $\offerfork$. Thus, the
scheduler context needs a communication channel for each cooperative
kernel. We implement the communication channels using atomic variables in global memory.

\paragraph{Launching kernels and managing workgroups}

To launch a kernel, the host sends a data packet to the GPU
consisting of: a kernel to execute, the inputs to the kernel, the
number of workgroups that should execute the kernel, and a flag indicating whether the kernel
is cooperative. In our implementation, this host-device communication channel is built using fine-grained SVM atomics.

On receiving a data packet from the host
describing a kernel launch $K$, the GPU scheduler must make decisions
about how to schedule $K$. Suppose $K$ requests
$N$ workgroups. The scheduler queries the scheduler context.  If
there are at least $N$ available workgroups, $K$ can be scheduled
immediately. Suppose instead that there are only $N_a < N$ available workgroups, but a cooperative kernel $K_c$ is
executing. The scheduler can use $K_c$'s channel in the scheduler context to command $K_c$ to provide $N - N_a$ workgroups via $\offerkill$.
Once $N$ workgroups are available, the scheduler initialises a
kernel context $\KC$ for $K$, setting the number of active workgroups
to $N$ and mapping available workgroups (located in the scheduler
context) to their new $\KC$ workgroup ids. The scheduler then sends $N$
workgroups from the available workgroups to execute kernel $K$ with
kernel context $\KC$.  Note that if the new kernel $K$ is itself a cooperative kernel, the scheduler could decide to provide $K$ with fewer than $N$ active workgroups initially.
